Cartogram Mapping and its Application to Cancer Data Visualization

### 2019-12-09 review_article.Rmd
language editing

Cancer atlases communicate cancer statistics over geographic domains. These domains are subdivided by administrative areas such as countries, states or suburbs. When communicating human-related statistics in Australia, the geographic map base highlights the sparsely populated rural areas. The smaller geographic areas may not draw the attention of readers but they are important to consider if they are more densely populated. Alternative map displays can reveal patterns that are not obvious when the same data is shown on a choropleth map. Alternative displays have proven to be successful on online news sites, especially for the communication of election data, where votes are aggregated at state level.

# Introduction
<!-- 436 words -->
<!-- What data are we working with, why, what are the purposes -->
Cancer statistics are often delivered as an aggregated value for a geopolitical area. Presenting these statistics requires transforming individual observations into aggregations of communities as geographical units, in large part for privacy protection or political and policy purposes.  The information could be as simple as counts per area (e.g. state, province, local government area, post/zip code). Counts alone are not sufficient to compare areas, because the populations of areas are all different. In this case, the count data needs to be merged with population data to appropriately calibrate it to incidence, for example, the rate per 100000 people.  This type of data is collected on a routine basis for public health purposes and may be made available to the general public as a service to the community. The task, then, is to examine what are the usual ways to communicate cancer statistics, to the public, are there alternative approaches, and what are the pros and cons of these choices.

<!-- Ways to visualize, whats our hypothesis -->
A common approach to communicate cancer statistics is to display statistics on a map. Using a choropleth map: the statistic is mapped to color and the geographic region is filled with this color. The viewer would then be able to examine the spatial distribution of the disease incidence, where there is a trend in longitude or latitude, or rural vs urban, or coastal vs inland, or even specific hot spots of the disease. Visualizing diseases on maps is often the first step in exploratory spatial data analysis and effectively helps in the formulation of hypotheses. Disease maps help to present geographic patterns that may be overlooked in a table, obscuring the geospatially related statistics [@SAMGIS].
By providing a visual representation of cancer outcomes, geographic patterns of the disease may be identified and effectively addressed with public health policy and actions.
Daniel Exeter [@SE] recognizes one of the key challenges with mapping spatial patterns of disease is the design of visualizations.
This paper addresses the visualization techniques and their applications to cancer statistics. Highlighting the differences and historic use of these displays.

<!-- Structure of the paper -->
The paper is organized as follows. The next section describes the choropleth map which is the common approach to disease maps. Section \ref{ch:public} surveys atlases in use today. Section \ref{ch:alternatives} describes  alternative displays, including the cartogram which is useful when the map has heterogeneously sized geographic units. The pros and cons of these approaches are discussed in Section \ref{ch:compare}. Disease maps are more useful when made interactive, and common options are described in Section \ref{ch:interacting}, along with a discussion of benefits and disadvantages. The last section summarizes the paper.

# Traditional approaches for cancer map displays {#ch:choropleths}
<!-- 381 words -->
<!-- what is a choropleth map -->
A choropleth map is used to display differences in the geographical distribution of data by spatial unit by shading areas of a map [@EI]. The geography is faithfully rendered, and the color rendering is designed to reveal spatial patterns among data values. Identifying and explaining spatial structures, patterns, and processes involve considering the individuals in communities and organizing communities into representable units [@SAMGIS].
Early versions of choropleth maps used symbols or patterns instead of color.
The use of choropleths to visualize cancer data [@CPISACA], and the development of these maps for displaying disease data [@DMAHP]. In epidemiology, choropleths are often used as a tool to study the spatial distribution of cancer incidence and mortality.

Figure \ref{fig:choroCRS} shows various geographic projections of choropleth maps of age-adjusted rate (per 100,000 people) of new cases of lung and bronchus in the USA, averaged from 2012 to 2016. The data was extracted from the official federal statistics on cancer incidence and deaths [@USInteractive], produced by the Centers for Disease Control and Prevention (CDC) and the National Cancer Institute (NCI). The rates exhibit a spatial trend, increasing from west to east. There is also a spatial outlier -- Utah has a noticeably lower rate than its neighbors. Also, Kentucky has a noticeably high rate, and Maine also has a higher rate than its neighbors. There is something of a cluster of higher rates around the tobacco states.

<!-- psychology -->
Utilizing the state boundaries can make a map familiar to read [@CIBMUK], and allows viewers to visually infer the spatial relationships in the data. 
The familiarity of the geography is a worthy consideration when presenting results of spatial analysis, to consumers that extend beyond researchers, to the public, policymakers and the media [@CPISACA].

<!-- potential problems -->
While the areas are recognizable shapes, they are often politically driven boundaries with individual areas being of non-uniform size, containing different population densities and subject to change over time. 
The different population and geographical sizes of administrative areas can attract attention to the shades of the underpopulated but large areas [@EI], Skowronnek [-@BCM] calls this an area-size bias. 
Choropleths can inhibit visual inference when presenting human-related statistics as the display may draw attention away from the 'potentially more important results in the more populous communities' that are geographically smaller [@SE].

@@

## Cancer atlases {#ch:public}
<!-- 360 words -->
<!-- Displaying of cancer data on choropleths -->
A cancer atlas is a map, or collection of maps, representing cancer incidence and mortality for a country, or group of countries. Atlases are key to developing hypotheses regarding areas with unusually high rates, geographic correlations, work-related exposures, and high-risk diets [@MACM]. The data collection of statistics across regions and the administrative control within regions lends itself to choropleth visualization. 
Atlases date to Haviland's maps in 1875, with UK cancer atlases appearing in 1961 [@HEDP], and early work in US cancer atlases appearing in 1971 [@burbank]. The presentation of cancer statistics has increased with greater access to computational power, and the availability of geographic information systems software [@SE].

<!-- Measures reported in cancer atlases -->
Mortality rates are now often presented as relative rates of risk across the population, and age-adjusted to correct for the higher prevalence of cancers in older populations.
Howe [-@HEDP] describes Stock's development of the standardized mortality ratios through the 1930s. Table \ref{tab:measures} summarizes the measures commonly presented in published cancer atlases.

@@

<!-- Cancer atlas literature -->

Cancer maps are effective tools for communicating incidence, survival, and mortality to a wide range of audiences, including the general public and others not trained in statistical analyses.
These visualizations enable non-expert audiences to interpret the outputs of sophisticated statistical analyses. Cruickshank (1947) as cited by S. D. Walter [@DMAHP], discusses using visuals as a 'formal statistical assessment of the spatial pattern'. Overwhelmingly, the cancer maps utilized to communicate to the public and other non-expert audiences are choropleths.

<!-- written below The Atlas of Cancer in Queensland [@DACQ] focused on highlighting the difference in experience for those living in rural and disadvantaged areas, the Standardized Incidence Ratios (SIR) were modeled via Bayesian methods.
The presentation of these rates considers not only the shapes of the areas, but also appropriate choices of color blind friendly color schemes, and categories of the values to highlight significantly different areas. 
-->

## Overview of publicly available atlases \label{sec:publicatlases}
<!-- 416 words -->
Roberts [@roberts2019communication] identified 33 publicly available cancer atlases, published between January 2010 and November 2015, from different countries around the globe. All of these use choropleth maps. All but one of these were published by non-commercial organizations, including not-for-profits, government, research organizations, advocacy groups or government-funded partnerships.  Figure \ref{fig:choropleth-grid-create} displays a subset of maps from these atlases.

<!-- While there is significant variation across these maps in terms of the measure represented, the scale of the map, the color choices, the legend design and supplementary/supporting statistics, the mapping approach has not changes and almost exclusively only choropleth maps are used. Unlike innovation in spatial statistical methods, interactivity, data collection and record keeping little work has been done to implement different mapping approaches -->

@@

With any number comes some uncertainty about its value. This can occur for several reasons: numbers are drawn from samples, errors occur in the disease reporting, values are simulated from a model to maintain privacy. Close to half of the atlases identified (42%, n=14) included some measure of uncertainty. Uncertainty is often recorded as sample or population size, standard deviation, confidence intervals, statistical significance, quantiles, credible intervals. The most common measure used to represent uncertainty were credible or confidence intervals (CIs).

@@

# Limitations of choropleth displays

Australia presents an extreme case of an urban-rural divide. The landmass occupied by urban electoral districts is only 10% of Australia, yet 90% of the population live in these urban areas. To present election results on a choropleth map should be 'unthinkable', as it means diminishing the visual impact of the majority of the electorates.
Dorling [@ACTUC] has a cartogram showing the 1966 general election results, which looked very little like the geographical shape of Australia. This issue is felt in any nation which experiences a spatially heterogeneous population distribution. As this feature of population distributions continues to intensify, the need for cartograms as an alternative to a choropleth map should only increase.

@@

# Contemporary alternatives to choropleths {#ch:alternatives}

@@

<!-- Denominators -->
The purposeful distortion of the map space, transformed according to population density, is beneficial when a uniform density of the map base is desired. The population then becomes a uniformly distributed background for the statistic presented [@GOINO]. "population distribution is often extremely uneven in former British colonies" [@ACTUC ], and this makes the distortion necessary [@CTTMB].
When implementing a distortion of the geographical shape according to population, the resulting display is an area cartogram [@NAC], or population-by-area cartogram [@TAAM]. Event cartograms change the area of regions on a map depending on the amount of disease-related events, but this does not consider the effects of land area and population [@VSSDCUC].

@@

A contiguous cartogram maintains connectivity of the map regions while areas are altered according to a statistic. This transformation often occurs at the expense of the shape of areas [@CBATCC, @NAC, @TAAM].
Three methods are provided for creating value-by-area cartograms [@ACA]. They implement 'map deformation' to account for the value assigned to each area.
Other methods include Tobler’s Pseudo-Cartogram Method, Dorling’s Cellular Automaton Method [-@ACTUC], Radial Expansion Method of Selvin et al., Rubber Sheet Method of Dougenik et al., Gusein-Zade and Tikunov’s Line Integral Method, Constraint-Based Method (Kocmoud and House) [-@CBATCC].

@@

Cartograms provide the spatial distortion to more accurately convey the statistical distribution, focuses on the human impact of the disease. However, if the population density is highly dissonant with geographic density the cartogram will lose all spatial context, and thus futile.

@@

The Australian Cancer Atlas [-@TACA] provides [tours](
https://atlas.cancer.org.au/app/tour/lungcancer) that change the display to the draw users' attention to areas on the map that are relevant to the story.
This implementation of animation gives users tools to plan their exploration.

@@

<!-- Animations can be implemented to allow users varying degrees of control.
Figure \ref{fig:animating} a gives an example of how time can be included through animating displays. This is a basic form of interactivity with animation allowing users to play, pause and rewind films, where the film is intended to be watched in one direction but the position in the animation can be controlled.
@CPISACA provide weather maps as a thoroughly developed example of animation of spatial displays to communicate information to the general public.
The movement of a weather system will follow a forecasted path, all map users can follow the animated path of the weather system across the geography over a specified period of time.
Where the message is most important, static or animated graphics allow control over the display and interpretation.
User interaction more allow clicks to progress between stages of more complex animations.
The Australian Cancer Atlas [-@TACA] provides [tours](
https://atlas.cancer.org.au/app/tour/lungcancer) which change the display to draw user's attention to areas on the map that are relevant to the story.
This implementation of animation gives users tools to plan their own exploration.
-->

>>>>

Cancer atlases communicate cancer statistics over geographic domains. These domains are subdivided into administrative regions such as countries, states or suburbs. When communicating human-related statistics in Australia, the geographic map base draws attention to the sparsely populated rural areas. The smaller geographic areas may not draw the attention of readers but they are important to consider if they are more densely populated. Alternative map displays can reveal patterns that are not obvious when the same data is shown on a choropleth map. Alternative displays have proven to be successful on online news sites, especially for the communication of election data, where votes are aggregated at state level.

# Introduction
<!-- 436 words -->
<!-- What data are we working with, why, what are the purposes -->
Cancer statistics are often delivered as aggregated values for geopolitical areas. Presenting these statistics requires aggregating individual observations into statistics for the geographical units, especially for privacy protection but also for political and policy purposes. Counts per area require minimal transformation (e.g. state, province, local government area, post/zip code). However, counts alone are not sufficient for the comparison areas, because the populations of areas are all different. The count data needs to be merged with population data to present incidence rates, for example, the rate per 100000 people. This type of data is collected on a routine basis for public health purposes and may be made available to the general public as a service to the community.

<!-- Ways to visualize, whats our hypothesis -->
Map displays are commonly required for the communication of cancer statistics over geographic domains. When using a choropleth map: the statistic is mapped to color and the geographic region is filled with this color. Disease maps help to present geographic patterns that may be overlooked in a table, obscuring the geospatially related statistics [@SAMGIS]. Providing a visual representation of cancer outcomes allows geographic patterns of the disease to be identified and effectively addressed with public health policy and actions. The spatial distribution of the disease incidence can be examined using a choropleth and may reveal a trend in longitude or latitude, or rural vs urban, or coastal vs inland, or even specific hot spots of the disease. Visualizing diseases on maps is often the first step in exploratory spatial data analysis and effectively helps in the formulation of hypotheses. 
Daniel Exeter [@SE] recognizes one of the key challenges with mapping spatial patterns of disease is the design of visualizations. This paper considers the current visualization techniques to communicate statistics to the public, and their applications to cancer statistics. Alternative approaches are also considered, and the limitations of the visualization methods, highlighting the differences and historic use of these displays.

<!-- Structure of the paper -->
The paper is organized as follows. The next section describes the choropleth map which is the common approach to disease maps and presents examples of atlases in use today. The limitations of the choropleth map are discussed in Section \ref{ch:chorolimit}. Section \ref{ch:alternatives} describes alternative displays, including the cartogram which is useful when the map has heterogeneously sized geographic units. Section \ref{ch:altlimits} presents the limitaions of producing and using alternative displays. Disease maps are more useful when made interactive, and common options are described in Section \ref{ch:interacting}, along with a discussion of benefits and disadvantages.

# Traditional approaches for cancer map displays {#ch:choropleths}
<!-- 381 words -->
<!-- what is a choropleth map -->
A choropleth map displays the geographical distribution of data over a set of spatial units by shading areas of a map [@EI]. The geography of a choropleth is faithfully rendered, and the color rendering is designed to reveal spatial patterns among data values. Identifying and explaining spatial structures, patterns, and processes involve considering the individuals and organizing them into representable units of communities [@SAMGIS].
Early versions of choropleth maps used symbols or patterns instead of color.
Choropleths can be used for displaying disease data [@DMAHP], including cancer data [@CPISACA]. In epidemiology, choropleths are often used as a tool to study the spatial distribution of cancer incidence and mortality.

Figure \ref{fig:choroCRS} shows various geographic projections of choropleth maps of age-adjusted rate (per 100,000 people) of new cases of lung and bronchus in the USA, averaged from 2012 to 2016. The data was extracted from the official federal statistics on cancer incidence and deaths [@USInteractive], produced by the Centers for Disease Control and Prevention (CDC) and the National Cancer Institute (NCI). The rates exhibit a spatial trend, increasing from west to east. There is also a spatial outlier -- Utah has a noticeably lower rate than its neighbors. Also, Kentucky and Maine have higher rates than their neighbors. There is possibly a cluster of higher rates around the tobacco states.

<!-- psychology -->
Displaying familiar state boundaries can make a map simpler to read [@CIBMUK], and allow viewers to visually infer the spatial relationships in the data using their mental model of the geography.
Consumers of disease displays may include researchers, the general public, policymakers and the media [@CPISACA], the familiarity of the geography is a worthy consideration when presenting results of spatial analysis.

<!-- potential problems -->
The familiar shapes are often politically driven boundaries with individual areas being of non-uniform size, containing different population densities and subject to change over time. 
The different population and geographical sizes of administrative areas can attract attention to the shades of the underpopulated but large areas [@EI], Skowronnek [@BCM] calls this an area-size bias. 
Choropleths can inhibit visual inference when presenting human-related statistics as the display may draw attention away from the 'potentially more important results in the more populous communities' that are geographically smaller [@SE].

@@

## Cancer atlases {#ch:public}
<!-- 360 words -->
<!-- Displaying of cancer data on choropleths -->
A cancer atlas is a map, or collection of maps, representing cancer incidence and mortality for a country, or group of countries. Atlases are key to developing hypotheses regarding areas with unusually high rates, geographic correlations, work-related exposures, and high-risk diets [@MACM]. The data collection methods across regions and the administrative control within regions lends itself to choropleth visualization. 
Cancer maps and atlases date back to Haviland's maps in 1875, and early work in US cancer atlases appearing in 1971 [@burbank]. The presentation of cancer statistics has increased with greater access to computational power, and the availability of geographic information systems software [@SE].

<!-- Measures reported in cancer atlases -->
Cancer maps are effective tools for communicating incidence, survival, and mortality to a wide range of audiences, including the general public and others not trained in statistical analyses.
These visualizations enable non-expert audiences to interpret the outputs of sophisticated statistical analyses. Cruickshank (1947) as cited by S. D. Walter [@DMAHP], discusses using visuals as a 'formal statistical assessment of the spatial pattern'. Overwhelmingly, the cancer maps utilized to communicate to the public and other non-expert audiences are choropleths.

The statistics presented have been developed over time. Mortality rates are now often presented as relative rates of risk across the population, and age-adjusted to correct for the higher prevalence of cancers in older populations.
Howe [@HEDP] describes Stock's development of the standardized mortality ratios through the 1930s. Table \ref{tab:measures} summarizes the measures commonly presented in published cancer atlases.

@@

## Overview of publicly available atlases \label{sec:publicatlases}
<!-- 416 words -->
Roberts [@roberts2019communication] identified 33 publicly available cancer atlases, published between 2010 and 2018, from different countries around the globe. All of these online atlases used choropleth maps. All but one of these were published by non-commercial organizations, including not-for-profits, government, research organizations, advocacy groups or government-funded partnerships.  Figure \ref{fig:choropleth-grid-create} displays a subset of maps from these atlases.

@@

Uncertainty about the value of a statistic is often communicated in cancer atlases. Uncertainty may occur for several reasons: numbers are drawn from samples, errors occur in the disease reporting, or values are simulated from a model to maintain privacy.  Uncertainty is often recorded as sample or population size, standard deviation, confidence intervals, statistical significance, quantiles, credible intervals. The most common measure used to represent uncertainty were credible or confidence intervals (CIs).

@@

# Limitations of choropleth displays {#ch:chorolimit}

A choropleth display may be limited by the administrative boundaries used to define regions.
This issue is often discussed in the context of election mapping, where each electorate decides the outcome of one parliamentary seat in the house or representatives.
While all states are visible in a map display of Australia, smaller regions such as electorates present an extreme case of an urban-rural divide. 
The landmass occupied by urban electoral districts is only 10% of Australia, yet 90% of the population live in these urban areas. To present election results on a choropleth map should be 'unthinkable', as it means diminishing the visual impact of the majority of the electorates.
Using a map of Australia as a visual statistic would communicate an incorrect distrbution of votes across the country. The issue of spatially heterogeneous population distribution across a country's landmass is common to many countries, especially British colonies [@ACTUC].

@@

 As this feature of population distributions continues to intensify, the need for cartograms as an alternative to a choropleth map should only increase.
# Contemporary alternatives to choropleths {#ch:alternatives}

@@

<!-- Denominators -->
The purposeful distortion of the map space, transformed according to population density, is beneficial when a uniform density of the map base is desired. The population then becomes a uniformly distributed background for the statistic presented [@GOINO], "population distribution is often extremely uneven in former British colonies" [@ACTUC], and this makes the distortion necessary [@CTTMB].
When implementing a distortion of the geographical shape according to population, the resulting display is an area cartogram [@NAC], or population-by-area cartogram [@TAAM]. Event cartograms change the area of regions on a map depending on the amount of disease-related events, but this does not consider the effects of land area and population [@VSSDCUC].

@@

A contiguous cartogram maintains connectivity of the map regions while areas are altered according to a statistic. This transformation often occurs at the expense of the shape of areas [@CBATCC, @NAC, @TAAM].
Three methods are provided for creating value-by-area cartograms [@ACA]. They implement 'map deformation' to account for the value assigned to each area.
Other methods include Tobler’s Pseudo-Cartogram Method, Dorling’s Cellular Automaton Method [@ACTUC], Radial Expansion Method of Selvin et al., Rubber Sheet Method of Dougenik et al., Gusein-Zade and Tikunov’s Line Integral Method, Constraint-Based Method (Kocmoud and House) [@CBATCC].

@@

Cartograms provide the spatial distortion to more accurately convey the statistical distribution, focuses on the human impact of the disease. However, if the population density is highly dissonant with geographic density the cartogram will lose all spatial context, and thus futile. 
Dorling [@ACTUC] has a cartogram showing the 1966 general election results, which looked very little like the geographical shape of Australia.

@@

The Australian Cancer Atlas [@TACA] provides [tours](
https://atlas.cancer.org.au/app/tour/lungcancer) that change the display to the draw users' attention to areas on the map that are relevant to the story.
This implementation of animation gives users tools to plan their exploration.

@@

<!-- Animations can be implemented to allow users varying degrees of control.
Figure \ref{fig:animating} a gives an example of how time can be included through animating displays. This is a basic form of interactivity with animation allowing users to play, pause and rewind films, where the film is intended to be watched in one direction but the position in the animation can be controlled.
@CPISACA provide weather maps as a thoroughly developed example of animation of spatial displays to communicate information to the general public.
The movement of a weather system will follow a forecasted path, all map users can follow the animated path of the weather system across the geography over a specified period of time.
Where the message is most important, static or animated graphics allow control over the display and interpretation.
User interaction more allow clicks to progress between stages of more complex animations.
The Australian Cancer Atlas [@TACA] provides [tours](
https://atlas.cancer.org.au/app/tour/lungcancer) which change the display to draw user's attention to areas on the map that are relevant to the story.
This implementation of animation gives users tools to plan their own exploration.
-->
